Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-11 paper/main.tex
final edits

\begin{equation}
	\tau = \sum_{\substack{X, Y \in V \\ X \rightarrow Y, Y \rightarrow X \not \in E}}   \phi(X, Y, \mathrm{pa}_G(X) \cup \mathrm{pa}_G(Y))
\end{equation}

@@

Another key difference lies in the interpretability of the evaluation metric.
Scoring metrics are based on usually based on the log-likelihood with a penalty
for model complexity. They can be used to make relative comparisons between
models but their values do not have any interpretation in an absolute sense.
That is, they indicate which model is better for a given dataset but do not
quantify how well the model explains the data in an absolute sense. In
contrast, $ \tau $ could be seen as an absolute measure of model fit -- 
its value approaches $ 0 $ as the model perfectly explains the observed data.

@@

In this section, we compare our \textsc{ExpertInLoop} algorithm with automated
causal discovery algorithms. Our goal with the empirical analysis is to
understand the behaviour of our method when the experts are not perfect. To
simulate an imperfect domain expert, we use a version of the strong ancestral oracle
that has an accuracy of $ \alpha $:
\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathcal{O}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
			\mathcal{O}(X, Y) &= \begin{cases}
				X \rightarrow Y, & \textrm{if } X \in \textrm{An}_{G_{\textrm{true}}}(Y) \\
				X \leftarrow Y, & \textrm{if } Y \in \textrm{An}_{G_{\textrm{true}}}(X) \\
				\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}
where $G_{\textrm{true}}$ is the true data generating DAG.

It is important to note that the effective accuracy of $ \mathrm{Expert} $ is
higher than $ \alpha $ as even when $ x > \alpha $, there is a $ 1 $ in
$ 3 $ chance that edge orientation is correct. Therefore, an $ \mathrm{Expert} $ with
accuracy $ \alpha $, has an effective accuracy, $ \alpha_{\mathrm{eff}} =
\alpha + (1 - \alpha) / 3 $.

@@

We analyzed the performance of \textsc{ExpertInLoop} algorithm by comparing it
with three other automated algorithms: PC, Hill-Climb Search, and Greedy
Equivalence Search (GES) on linear Gaussian data. For \textsc{ExpertInLoop}, we
use the \textsc{RankedExpand} method with Pearson's correlation coefficient as
the effect size (Section~\ref{sec:ci_tests}) and significance threshold
$\alpha=0.05$; note that this means that also the d-separation ``oracle''
makes both type~I and type~II errors. For the analysis, we
simulate data from known DAGs and use the algorithms to recover the original
DAG. We start by generating a random DAG on $10$ nodes and use linear models
with random effects to simulate $ 500 $ samples from it. We simulate each
variable $ V_i $ in the DAG as
$$
	V_i = \bm{\beta} \cdot \mathrm{Pa}_G(V_i) + \mathcal{N}(0, 1)
$$
where each coefficient $\beta_i$ of the vector $\bm{\beta} \in \mathbb{R}^{|\bm{\beta} |}$ is randomly drawn as
$$
\beta_i \sim \mathrm{Uniform}([-0.6, 0.6])
$$

We perform the experiment for both varying densities of the DAG and accuracy of
$\mathrm{Expert}$. We also repeat each experiment $ 30 $ times and report the
mean and standard error of the results.

To compare the learned DAG to the original DAG, we use two metrics: Structural
Hamming Distance (SHD) and Structural Intervention Distance (SID)
\citep{Peters2015}. Unlike \textrm{ExpertInLoop}, the automated algorithms can
only recover the MEC, therefore, we considered all possible orientations of the
MEC to compute a range of SHD and SID values representing the best and worst
case scenarios. The results of this analysis are shown in
Figure~\ref{fig:shd_sid}. For SHD, the performance of \textsc{ExpertInLoop} is
comparable to the automated algorithms for $ \alpha = 0.3
(\alpha_{\textrm{eff}} = 0.53 $), implying that if an expert is able to get the
edge orientation correctly in more than $ 1 $ out of $ 2 $ cases, they can
outperform the automated algorithms. Similarly for SID, an expert can outperform 
the automated algorithms if they are able to get the orientation correct in $ 2 $ 
out of $ 3 $ cases. Another thing to note is that for denser DAGs, the 
\textsc{ExpertInLoop} is able to perform better for lower $ \alpha $ values.

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{LLM-in-the-loop causal discovery on the Adult Income datases. As the ground 
truth is unknown, two measures of fit are shown over 30 iterative modifications: 
total residual association $\tau$ (a, lower is bettter); log-likelihood (a, higher is better).
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

We now present two proof-of-concept implementations of our approach, 
geared towards using  large language models (LLMs) and humans as experts
to decide on edge directions. We use the Adult Income dataset from the 
introduction as an example. There is no known ground truth for this dataset,
and it contains mixed types of data; here, we use a version that contains
ordinal and nominal variables and use the RMSEA as a measure of association
to prioritize modifications (Section~\ref{sec:background}).

@@

\begin{figure}[t!]
	\centering
	\includegraphics[page=1]{fig5.pdf}
	\caption{DAG learned from the Adult Income dataset using an LLM
		(Gemini 1.5 Flash) as the expert. The p-value threshold used is $ 0.05 $ 
		and the measure of association threshold is $ 0. 1 $}
	\label{fig:adult_llm}
\end{figure}

@@

Since our approach relies on expert knowledge to determine edge orientations,
we explored the potential of using an LLM for this task. We applied our causal
discovery procedure using the Gemini 1.5 Flash model as the expert. We simulate
the \emph{weak ancestral oracle} using the LLM by asking it to choose the
causal direction between a pair of variables. The LLM is provided a description
of each of the variables and given the prompt shown in
Appendix~\ref{section:llms}. 
Figure~\ref{fig:unexplained_ll} shows how the model fit to the data improves 
across 30 iterations compared to GES for a run of the algorithm.
Figure~\ref{fig:adult_llm} shows the DAG learned
on the Adult Income dataset using the LLM as the expert. We can see 
that the model fit improves comparably to a greedy approach and 
the resulting DAG contains largely sensible edge directions.

@@

To enable researchers to easily apply our approach to their own datasets, we
developed an interactive web-based tool (Figure~\ref{fig:web}) for constructing
DAGs. Users can upload their dataset, which initializes an empty DAG with nodes
corresponding to the dataset's variables. They can then specify a p-value
threshold and a threshold for a minimal association strength. The tool 
then visually highlights variable pairs
with a residual association greater than the threshold, marking them with red
edges; to avoid cluttering, no lines are drawn for residual associations that
are statistically insignificanct or below the specified threshold.
The thickness of these edges represents the strength of association,
helping users prioritize which variable pairs to address first; however, note that
users are free to choose which variables to connect. Similarly, if
an existing edge is found to be potentially superfluous (statistically insignificant),
it is highlighted in black. Using this information, users can iteratively
modify the model by adding or removing edges. The tool also computes the Root
Mean Square Error of Approximation (RMSEA) based on Shipley's C test, a global test 
of model fit based on the implied CIs \citep{Shipley2000}, to provide an estimate of the
overall fit of the model. Once satisfied with the constructed DAG, users can
export the model for further analysis.

@@

Researchers often prefer to construct DAGs manually rather than using automated
algorithms due to practical challenges in applying automated methods. To assist
this process, we developed an iterative structure learning method that
integrates manual construction with data-driven feedback, bridging
the gap between fully manually and fully automated methods.
The idea of augmenting structure learning by domain knowledge is of course not 
new. Common ways to provide such background knowledge are "whitelists" or "blacklists"
of edges 
\citep{scutari2010}, or specification
of a temporal structure between the variables \cite{Bang2023}.
Compared to such approaches, where domain knowledge 
is specified and supplied beforehand, the novelty 
of our method lies in the interactive
back-and-forth between domain knowledge and data, 
which provably leads to the correct
result in a polynomial amount of steps (if the expert answers all
ancestral queries correctly).

>>>>

\begin{equation}
	\tau = \sum_{\substack{X, Y \in V \\ X \rightarrow Y, Y \rightarrow X \not \in E}}   \phi(X, Y, \mathrm{pa}_G(X) \cup \mathrm{pa}_G(Y))
\label{eqn:tau}
\end{equation}

@@

Another key difference lies in the interpretability of the evaluation metric.
Scoring metrics are based on usually based on the log-likelihood with a penalty
for model complexity. They can be used to make relative comparisons between
models but their values do not have any interpretation in an absolute sense.
That is, they indicate which model is better for a given dataset but do not
quantify how well the model explains the data. In
contrast, $ \tau $ could be seen as an absolute measure of model fit -- 
its value approaches $ 0 $ as the model perfectly explains the observed data.

@@

In this section, we compare our \textsc{ExpertInLoop} algorithm with automated
causal discovery algorithms. Our goal with the empirical analysis is to
understand the behaviour of our method when both experts are not perfect.
Specifically, how good does the domain expert have to be so we actually benefit
from their expertise, compared to a purely data-driven approach?

In our analysis, we simulate data from a ``true'' DAG $G$ and use the
{\sc ExpertInLoop} algorithm with the {\sc RankedExpand} heuristic 
to recover $G$. We implement the $d$-separation oracle $\mathcal{D}_G(X,Y,\mathbf{Z})$
by conducting conditional independence tests of $X \ci Y \mid \mathbf{Z}$, which on finite
data inevitably make type~I and type~II errors. To
simulate an imperfect domain expert, we use a version of a strong 
ancestral oracle $\mathcal{A}_G$ (Section~\ref{sec:modification})
that with probability $\alpha$ knows the correct answer and randomly
guesses otherwise, i.e.,
\begin{equation*}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathcal{A}_G(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} 
	\end{split}
\end{equation*}

It is important to note that the effective accuracy of $ \mathrm{Expert} $ is
higher than $ \alpha $ as even when $ x > \alpha $, there is a $ 1/3 $
that edge orientation is correct. Therefore, an $ \mathrm{Expert} $ with
accuracy $ \alpha $, has an effective accuracy, $ \alpha_{\mathrm{eff}} =
\alpha + (1 - \alpha) / 3 $.

@@

We compare the performance of the \textsc{ExpertInLoop} algorithm by
to three other automated algorithms: PC \citep{Spirtes2001,KalischB07}, 
Hill-Climb Search \citep{scutari2010}, and Greedy Equivalence Search (GES) \citep{Chickering2002}
on linear-Gaussian data. For \textsc{ExpertInLoop}, we
use the \textsc{RankedExpand} method with Pearson's correlation coefficient as
the effect size (Section~\ref{sec:ci_tests}) and significance threshold
(and type~I error rate) $\alpha=0.05$. We start by generating a random DAG 
on $10$ nodes and use linear models
with random effects to simulate $ 500 $ samples from it. We simulate each
variable $ V_i $ in the DAG as
$$
	V_i = \bm{\beta} \cdot \mathrm{Pa}_G(V_i) + \mathcal{N}(0, 1)
$$
where each coefficient $\beta_i$ of the vector $\bm{\beta} \in \mathbb{R}^{|\mathrm{Pa}_G(V_i)|}$ is randomly drawn as
$$
\beta_i \sim \mathrm{Uniform}([-0.6, 0.6])
$$

We perform the experiment for varying densities of the DAG and accuracies of
$\mathrm{Expert}$. We repeat each experiment $ 30 $ times and report the
mean and standard error of the results.

To compare the learned DAG to the original DAG, we use two metrics: the Structural
Hamming Distance (SHD) and the Structural Intervention Distance (SID)
\citep{Peters2015}. Unlike \textsc{ExpertInLoop}, the automated algorithms can
only recover the MEC, therefore, we considered all possible orientations of the
MEC to compute a range of SHD and SID values representing the best and worst
case scenarios. The results of this analysis are shown in
Figure~\ref{fig:shd_sid}. For SHD, the performance of \textsc{ExpertInLoop} is
comparable to the automated algorithms for $ \alpha = 0.3$
($\alpha_{\textrm{eff}} = 0.53 $), implying that if an expert is able to get the
edge orientation correctly in more than $ 1 $ out of $ 2 $ cases, they can
outperform the automated algorithms. Similarly for SID, an expert can outperform 
the automated algorithms if they are able to get the orientation correct in $ 2 $ 
out of $ 3 $ cases. Another thing to note is that for denser DAGs, the 
\textsc{ExpertInLoop} is able to perform better for lower $ \alpha $ values.

\begin{figure}[t!]
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/unexplained_effect.pdf}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{0.25\textwidth}
		\centering
		\includegraphics{../code/fig4/ll.pdf}
		\caption{}
	\end{subfigure}
	\caption{LLM-in-the-loop causal discovery on the Adult Income datases. As the ground 
truth is unknown, two measures of fit are shown over 30 iterative modifications: 
total residual association $\tau$ (a, see Equation~\ref{eqn:tau}; lower is bettter); log-likelihood (b, higher is better).
	}
	\label{fig:unexplained_ll}
\end{figure}

@@

We now present two proof-of-concept implementations of our approach, 
geared towards using  large language models (LLMs) and humans as experts
to decide on edge directions. We use the Adult Income dataset from the 
introduction as an example. There is no known ground truth for this dataset,
and it contains mixed types of data; here, we use a version that contains
ordinal and nominal variables and use the RMSEA of a residualization-based
conditional independence test \citep{Ankan2023} as a measure of association
to prioritize modifications (Section~\ref{sec:background}).

@@

\begin{figure}[t!]
	\centering
	\includegraphics[page=1]{fig5.pdf}
	\caption{DAG learned from the Adult Income dataset using an LLM
		(Gemini 1.5 Flash) as the expert. The p-value threshold used is $ 0.05 $ 
		and the measure of association threshold is $ 0. 1 $, meaning
		that associations below $0.1$ are not considered by 
		\textsc{RankedExpand}.}
	\label{fig:adult_llm}
\end{figure}

@@

Since our approach relies on expert knowledge to determine ancestral relationships,
we explored the potential of using an LLM for this task. We applied our causal
discovery procedure using the Gemini 1.5 Flash model as the expert. We simulate
the \emph{weak ancestral oracle} using the LLM by asking it to choose the
causal direction between a pair of variables. The LLM is provided a description
of each of the variables and given the prompt shown in
Appendix~\ref{section:llms}. 
Figure~\ref{fig:unexplained_ll} shows how the model fit to the data improves 
across 30 iterations compared to GES for a run of the algorithm.
Figure~\ref{fig:adult_llm} shows the DAG learned
on the Adult Income dataset using the LLM as the expert. We can see 
that the model fit improves comparably to a greedy approach and 
the resulting DAG appears to contain largely sensible edge directions.

@@

To enable researchers to easily apply our approach to their own datasets, we
developed an interactive web-based tool (Figure~\ref{fig:web}) for constructing
DAGs. Users can upload their dataset, which initializes an empty DAG with nodes
corresponding to the dataset's variables. They can then specify a p-value
threshold and a threshold for a minimal association strength. The tool 
then visually highlights variable pairs
with a residual association greater than the threshold, marking them with red
edges; to avoid cluttering, no lines are drawn for residual associations that
are statistically insignificant or below the specified threshold.
The thickness of these edges represents the strength of association,
helping users prioritize which variable pairs to address first; however, note that
users are free to choose which variables to connect. Similarly, if
an existing edge is found to be potentially superfluous (statistically insignificant),
it is highlighted in black. Using this information, users can iteratively
modify the model by adding or removing edges. The tool also computes the Root
Mean Square Error of Approximation (RMSEA) based on Shipley's C test, a global test 
of model fit based on the implied CIs \citep{Shipley2000}, to provide an estimate of the
overall fit of the model. Once satisfied with the constructed DAG, users can
export the model for further analysis.

@@

Researchers often prefer to construct DAGs manually rather than using automated
algorithms due to practical challenges in applying automated methods. To assist
this process, we developed an iterative structure learning method that
integrates manual construction with data-driven feedback, bridging
the gap between fully manual and fully automated methods.
The idea of augmenting structure learning by domain knowledge is of course not 
new. Common ways to provide such background knowledge are ``whitelists'' or 
``blacklists'' of edges 
\citep{scutari2010}, or specification
of a ``tiered'' (e.g., temporal) structure between the variables \citep{Bang2023}.
Compared to such approaches, where domain knowledge 
is specified and supplied beforehand, the novelty 
of our method lies in the interactive
back-and-forth between domain knowledge and data, 
which provably leads to the correct
result in a polynomial amount of steps in the oracle settings we considered.
